### 2016-03-12 paper/icfp16-submission/semantics.tex
add proof of lem:resolve-compat
clean up

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon.
%
Thus, we must avoid the trap of generating \emph{spurious}
witnesses that arise from picking irrelevant values, when
instead there exist perfectly good values of a \emph{different}
type, under which the program would not have gone wrong.
%
We now show that our evaluation relation instantiates holes
in a \emph{general} manner. That is, given a function $f$,
if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then \emph{for every} input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any function $f$, if\\
  \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
  then for every % inhabitable
  type
  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
  $t$ there exists a $v$ of type $t$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{thm}

@@

\paragraph{The Type of a a Value} The \emph{dynamic type}
of a value $v$ is defined as a function $\typeof{v}$ shown
in Figure~\ref{fig:typeof}.
%
The types of primitive values are defined in the natural manner.
%
The types of functions are \emph{approximated}, which is all
that is needed to ensure an application does not get stuck, \eg
$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
instead of $\tint \rightarrow \tint$.
%
The types of (polymorphic) data types are obtained from the
labels on their values.

@@

\paragraph{Compatibility}
%
A \emph{type} $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t},
if $\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.
%
That is, two types are compatible if there exists a type substitution
that maps both types to the same type.
%
A \emph{value} $v$ is \emph{compatible} with a type $t$, written \vcompat{v}{t},
if $\tcompat{\typeof{v}}{t}$, that is, if the dynamic type of $v$ is
compatible with $t$.

@@

\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$, then
  $\vsub{\resolve{\thole}{\tsu'}}{\resolve{\ehole}{\vsu'}}$.
  %
  \ES{fill in cases in the appendix}
\end{proof}

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\thole$ does not change, so if the resolved types differ then
  $\tsu \neq \tsu'$.
  %
  Note that only \forcesym can change \tsu, and it can only do so
  via \unifysym, which can only extend \tsu.
  % and only the first case,
  % when we extend \vsu\ with $[\ehole \mapsto v]$,
  % does.
  %
  \ES{fill in cases in the appendix}
\end{proof}

@@

\begin{lem}
\label{lem:k-stuck}
For all $t$,
if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
   $\tincompat{t}{\ptype{\trace}{f}}$,
   then there exists a $v$ such that $\hastype{v}{t}$ and
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
   $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  We can construct $v$ from $\trace$ as follows.
  %
  Let
  $$
  \trace_i = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},
             \ldots,
             \triple{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}},
             \triple{e_{i}}{\vsu_{i}}{\tsu_{i}}
  $$
  be the shortest prefix of $\trace$ such that
  $\tincompat{\ptype{\trace_i}{f}}{t}$.
  %
  We will show that $\resolve{\thole}{\tsu_{i-1}}$
  must contain some other hole $\thole'$ that is
  instantiated at step $i$.
  %
  Furthermore, $\thole'$ is instantiated in such a way that
  $\tincompat{\resolve{\thole}{\tsu_{i}}}{t}$.
  %
  Finally, we will show that if we had instantiated $\thole'$ such that
  $\tcompat{\resolve{\thole}{\tsu_{i}}}{t}$, the current
  step would have gotten $\stuck$.

  % By Lemma~\ref{lem:fixme} we know that
  % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
  %
  By Lemma~\ref{lem:vsu-ext} we know that
  $\tsu_{i} = \tsu_{i-1}[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
  %
  We will assume, without loss of generality, that
  $\tsu_{i} = \tsu_{i-1}[\thole' \mapsto t']$.

  Next, we introduce a notion of type % and value
  contexts, defined
  analogously to evaluation contexts.
  %
  $$
  \begin{array}{lcl}
    T &::=& \bullet \spmid \ttree{T} \\
    % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
  \end{array}
  $$
  %
  Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
  types differ,
  $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
  and
  $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
  %
  Let $s$ be a concrete type such that $\intctx{s} = t$.
  %
  We show by case analysis on the evaluation rules that
  %
  $$\step{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}[\thole' \mapsto s]}{\stuck}{\vsu}{\tsu}$$
  %
  Finally, by Lemma~\ref{lem:resolve-compat} we know that
  $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
  thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
  %
  Let $u = \gen{s}{\tsu}$
  and $v = \resolve{\vhole{\thole}}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.
  %and
  % $\vsub{\resolve{\thole}{\tsu_{i}}}{\resolve{\ehole}{\vsu_{i}}}$.
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
  % =
  % \tcompat{\ptype{i-1}{f}}{t}
  % \\
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
  % =
  % \tincompat{\ptype{i}{f}}{t}
  %
  % Finally, by the definition of $\typeofsym$,
  % we know that $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{v}$
  % such that $\vsub{T'[\thole']}{v}$.
  %such that $\resolve{\typeof{v}}{\tsu_{i-1}} = T'[\thole']$
  %
  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  % Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  % we show by case analysis on the evaluation relation that
  % $\invctx{u}$ could not make progress at this step, and thus
  % $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}

>>>>

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon.
%
Thus, we must avoid the trap of generating \emph{spurious}
witnesses that arise from picking irrelevant values, when
instead there exist perfectly good values of a \emph{different}
type, under which the program would not have gone wrong.
%
We now show that our evaluation relation instantiates holes
in a \emph{general} manner. That is, given a lambda-term $f$,
if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then \emph{for every} input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any lambda-term $f$, if
  \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
  then for every % inhabitable
  type
  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
  $t$ there exists a $v$ of type $t$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{thm}

@@

\paragraph{The Type of a Value} The \emph{dynamic type}
of a value $v$ is defined as a function $\typeof{v}$ shown
in Figure~\ref{fig:typeof}.
%
The types of primitive values are defined in the natural manner.
%
The types of functions are \emph{approximated}, which is all
that is needed to ensure an application does not get stuck, \eg
$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
instead of $\tint \rightarrow \tint$.
%
The types of (polymorphic) data types are obtained from the
labels on their values.

@@

\paragraph{Compatibility}
%
A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t},
if $\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.
%
That is, two types are compatible if there exists a type substitution
that maps both types to the same type.
%
A value $v$ is \emph{compatible} with a type $t$, written \vcompat{v}{t},
if $\tcompat{\typeof{v}}{t}$, that is, if the dynamic type of $v$ is
compatible with $t$.

@@

\paragraph{Type Refinement}
A type $s$ is a \emph{refinement} of a type $t$, written $\tsub{s}{t}$,
if $\exists \theta. s = \subst{\theta}{t}$.
%
In other words, $s$ is a refinement of $t$ if there exists a type
substitution that maps $t$ directly to $s$.
%
A type $t$ is a \emph{refinement} of a value $v$, written $\tsub{t}{v}$,
if $\tsub{t}{\typeof{v}}$, \ie if $t$ is a refinement of the
dynamic type of $v$.
%
\ES{calling a type a refinement of a value sounds really weird..}

We prove that the partial instantiation of a type hole $\thole$
is always a refinement of the partial instantiation of the
associated value hole $\ehole$.

\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$, then
  $\vsub{\resolve{\thole}{\tsu'}}{\resolve{\ehole}{\vsu'}}$.
  %
\end{proof}

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$
% and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\thole$ does not change, so if the partial input types differ then
  $\tsu \neq \tsu'$.
  %
  Note that only \forcesym can change \tsu, and it can only do so
  via \unifysym, which can only extend \tsu.
  % and only the first case,
  % when we extend \vsu\ with $[\ehole \mapsto v]$,
  % does.
  %
  \ES{fill in cases in the appendix}
\end{proof}

@@

\begin{lem}
\label{lem:k-stuck}
For all types $t$,
if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
   $\tincompat{t}{\ptype{\trace}{f}}$,
   then there exists a $v$ such that $\hastype{v}{t}$ and
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
   $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  We can construct $v$ from $\trace$ as follows.
  %
  Let
  $$
  \trace_i = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},
             \ldots,
             \triple{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}},
             \triple{e_{i}}{\vsu_{i}}{\tsu_{i}}
  $$
  be the shortest prefix of $\trace$ such that
  $\tincompat{\ptype{\trace_i}{f}}{t}$.
  %
  We will show that $\ptype{\trace_{i-1}}{f}$ % $\resolve{\thole}{\tsu_{i-1}}$
  must contain some other hole $\thole'$ that is
  instantiated at step $i$.
  %
  Furthermore, $\thole'$ is instantiated in such a way that
  $\tincompat{\ptype{\trace_i}{f}}{t}$.
  % $\tincompat{\resolve{\thole}{\tsu_{i}}}{t}$.
  %
  Finally, we will show that if we had instantiated $\thole'$ such that
  $\tcompat{\ptype{\trace_i}{f}}{t}$,
  % $\tcompat{\resolve{\thole}{\tsu_{i}}}{t}$,
  the current step would have gotten $\stuck$.

  % By Lemma~\ref{lem:fixme} we know that
  % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
  %
  By Lemma~\ref{lem:vsu-ext} we know that
  $\tsu_{i} = \tsu_{i-1}[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
  %
  We will assume, without loss of generality, that
  $\tsu_{i} = \tsu_{i-1}[\thole' \mapsto t']$.
  %
  Next, we introduce a notion of type % and value
  contexts, defined
  analogously to evaluation contexts.
  %
  $$
  \begin{array}{lcl}
    T &::=& \bullet \spmid \ttree{T} \\
    % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
  \end{array}
  $$
  %
  Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
  types differ,
  $\ptype{\trace_{i-1}}{f} = \intctx{\thole'}$
  and
  $\ptype{\trace_i}{f} = \intctx{t'}$.
  % $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
  % and
  % $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
  %
  Let $s$ be a concrete type such that $\intctx{s} = t$.
  %
  We show by case analysis on the evaluation rules that
  %
  $$\step{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}[\thole' \mapsto s]}{\stuck}{\vsu}{\tsu}$$
  %
  Finally, by Lemma~\ref{lem:resolve-compat} we know that
  % $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
  $\vsub{\ptype{\trace_{i-1}}{f}}{\resolve{\ehole}{\vsu_{i-1}}}$,
  thus $\vsub{\intctx{\thole'}}{\resolve{\ehole}{\vsu_{i-1}}}$.
  %
  Let $u = \gen{s}{\tsu}$
  and $v = \resolve{\ehole}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.
  %and
  % $\vsub{\resolve{\thole}{\tsu_{i}}}{\resolve{\ehole}{\vsu_{i}}}$.
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
  % =
  % \tcompat{\ptype{i-1}{f}}{t}
  % \\
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
  % =
  % \tincompat{\ptype{i}{f}}{t}
  %
  % Finally, by the definition of $\typeofsym$,
  % we know that $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{v}$
  % such that $\vsub{T'[\thole']}{v}$.
  %such that $\resolve{\typeof{v}}{\tsu_{i-1}} = T'[\thole']$
  %
  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  % Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  % we show by case analysis on the evaluation relation that
  % $\invctx{u}$ could not make progress at this step, and thus
  % $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}
